---
title: "Data Dictionary"
vignette: >
  %\VignetteIndexEntry{data-dictionary}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    warning: false
    message: false
---

```{r}
#| label: setup
#| include: false
library(metrosp)
library(sf)
```

This vignette documents the core datasets shipped with `metrosp`: what each
contains, where the data comes from, and the caveats you should know before
analysing it. For auxiliary lookup tables (`metro_lines`, `metro_colors`,
`station_inauguration`) and derived datasets (`forecasts`,
`forecast_accuracy`), see the help pages (e.g., `?forecasts`).

# Overview

The six core datasets come from three public sources:

| Dataset | Grain | Source | Time span |
|---|---|---|---|
| `passengers_entrance` | line $\times$ month $\times$ metric | METRO + Dataverse | 2012--2025 |
| `passengers_transported` | line $\times$ month $\times$ metric | METRO | 2017--2025 |
| `station_averages` | station $\times$ month | METRO + Dataverse | 2012--2025 |
| `station_daily` | station $\times$ day | METRO + Dataverse | 2012--2025 |
| `lines` | line (spatial) | GeoSampa | current snapshot |
| `stations` | station (spatial) | GeoSampa | current snapshot |

: {.striped .hover}

All passenger counts are in **individual passengers** (not thousands). The
raw METRO portal files report values in thousands (*milhares*); the ETL
pipeline converts them to match the Dataverse source.

# Data sources

## METRO SP transparency portal {#source-metro}

The Companhia do Metropolitano de São Paulo (a.k.a. METRÔ) publishes monthly
demand reports at its [data transparency
portal](https://transparencia.metrosp.com.br/dataset/demanda). Reports
cover Lines 1 (Azul/Blue), 2 (Verde/Green), 3 (Vermelha/Red), 5
(Lilás/Lilac, until Jul 2018), and 15 (Prata/Silver), and are available from
October 2017 onward. Values are reported in thousands (*milhares*).

Three types of reports feed into the package datasets:

1. **Passageiros Entrada por Linha** --- monthly passenger *entries* by
   line, broken down by day-type metric.
2. **Passageiros Transportados por Linha** --- monthly passengers
   *transported* by line, broken down by day-type metric.
3. **Entrada de Passageiros por Estação - Média dos Dias Úteis** --- average
   weekday entries per station, per month.

Daily station-level data (one row per station per day) is available from 2020
onward.

Each monthly report breaks demand into five day-type metrics: total (monthly
aggregate), average on business days, average on Saturdays, average on
Sundays, and daily peak (maximum within the month).

## Insper Dataverse {#source-dataverse}

Lines 4 (Amarela/Yellow, operated by ViaQuatro) and 5 (Lilás/Lilac, operated
by ViaMobilidade from August 2018) are not published on the METRO portal.
Ridership data for these lines comes from the [Insper
Dataverse](https://doi.org/10.60873/FK2/UTGQ0I), starting January 2012
(Line 4) and August 2018 (Line 5). Transported counts are **not available**
for Lines 4 or 5.

Unlike the METRO data, Dataverse counts are not rounded to the nearest
thousand. For consistency, METRO values are multiplied by 1,000 during the
ETL so that all datasets report individual passengers.

The `station_averages` dataset for Lines 4 and 5 is derived from
`station_daily` using the `bizdays` package. Specifically, we use the
"Brazil/ANBIMA" calendar, which tracks days when the B3 stock exchange
operates in São Paulo. This closely mirrors the city's business-day schedule;
however, since 2022 B3 only closes for national holidays (not municipal or
state holidays such as the 9th of July). A more precise São Paulo business-day
calendar is planned for a future release.

## GeoSampa {#source-geosampa}

Spatial geometries for metro and commuter train (CPTM) lines and stations
come from [GeoSampa](https://geosampa.prefeitura.sp.gov.br/), the City of
São Paulo's open geospatial platform. The data includes both currently
operating infrastructure and planned future expansions.

# Passenger datasets

## passengers_entrance

Monthly passenger entries aggregated by metro line and day-type metric.

```{r}
dplyr::glimpse(passengers_entrance)
```

| Column | Type | Description |
|---|---|---|
| `date` | Date | First day of the month |
| `line_number` | integer | Line identifier (1, 2, 3, 4, 5, 15, or 99 for network total) |
| `metric_abb` | character | Metric code: `total`, `mdu`, `msa`, `mdo`, `max` |
| `value` | numeric | Passenger count |
| `metric` | character | Metric label in English |
| `metric_pt` | character | Metric label in Portuguese |
| `line_name` | character | Line color in English |
| `line_name_pt` | character | Line color in Portuguese |
| `year` | integer | Calendar year |

: {.striped .hover}

**Metrics:**

| Code | English | Portuguese |
|---|---|---|
| `total` | Total passengers in the month | Total |
| `mdu` | Average on business days | Média dos Dias Úteis |
| `msa` | Average on Saturdays | Média dos Sábados |
| `mdo` | Average on Sundays | Média dos Domingos |
| `max` | Daily peak | Máxima Diária |

: {.striped .hover}

**Coverage by line:**

| Line | Source | From | To |
|---|---|---|---|
| 1 -- Blue | METRO portal | Oct 2017 | present |
| 2 -- Green | METRO portal | Oct 2017 | present |
| 3 -- Red | METRO portal | Oct 2017 | present |
| 4 -- Yellow | Dataverse | Jan 2012 | present |
| 5 -- Lilac | METRO (Oct 2017--Jul 2018), Dataverse (Aug 2018+) | Oct 2017 | present |
| 15 -- Silver | METRO portal | Oct 2017 | present |
| 99 -- System | METRO portal | Oct 2017 | present |

: {.striped .hover}

## passengers_transported

Monthly passengers transported, aggregated by metro line and day-type
metric. Same structure as `passengers_entrance`.

```{r}
dplyr::glimpse(passengers_transported)
```

**Coverage by line:**

| Line | Source | From | To |
|---|---|---|---|
| 1 -- Blue | METRO portal | Oct 2017 | present |
| 2 -- Green | METRO portal | Oct 2017 | present |
| 3 -- Red | METRO portal | Oct 2017 | present |
| 5 -- Lilac | METRO portal | Oct 2017 | Dec 2019 |
| 15 -- Silver | METRO portal | Oct 2017 | present |
| 99 -- System | METRO portal | Oct 2017 | present |

: {.striped .hover}

Line 4 is absent entirely. The Dataverse source does not include transported
counts for Lines 4 or 5.

## station_averages

Monthly average weekday passenger entries per station.

```{r}
dplyr::glimpse(station_averages)
```

| Column | Type | Description |
|---|---|---|
| `date` | Date | First day of the month |
| `line_number` | integer | Line identifier |
| `station_name` | character | Full station name |
| `avg_passenger` | numeric | Average weekday (business day) entries |
| `line_name` | character | Line color in English |
| `line_name_pt` | character | Line color in Portuguese |
| `year` | integer | Calendar year |

: {.striped .hover}

Only the weekday average metric is available at the station level. For
line-level data with all five metrics, see `passengers_entrance`.

## station_daily

Daily passenger entries at each station.

```{r}
dplyr::glimpse(station_daily)
```

| Column | Type | Description |
|---|---|---|
| `date` | Date | Date of observation |
| `line_number` | integer | Line identifier |
| `station_name` | character | Full station name |
| `passengers` | numeric | Daily passenger entries |
| `line_name` | character | Line color in English |
| `line_name_pt` | character | Line color in Portuguese |
| `station_code` | character | Three-letter METRO abbreviation (`NA` for Lines 4--5) |
| `year` | integer | Calendar year |

: {.striped .hover}

**Coverage:**

- Lines 1, 2, 3, 15: 2020--present (METRO portal)
- Line 4: Jan 2012--present (Dataverse)
- Line 5: Aug 2018--present (Dataverse)

# Spatial datasets

The `lines` and `stations` datasets are `sf` objects in WGS 84 (EPSG:4326),
sourced from [GeoSampa](#source-geosampa). Both include currently operating
and planned future infrastructure for METRO SP and CPTM.

## lines

```{r}
dplyr::glimpse(lines)
```

| Column | Type | Description |
|---|---|---|
| `line_number` | integer | Official line number |
| `line_name_pt` | character | Line color in Portuguese |
| `line_name` | character | Line color in English |
| `company_name` | character | Operator (Metrô, ViaQuatro, ViaMobilidade, CPTM) |
| `type` | character | `"metro"` (underground) or `"train"` (CPTM commuter rail) |
| `status` | character | `"current"` (operating) or `"future"` (planned) |
| `geometry` | LINESTRING | Route geometry |

: {.striped .hover}

## stations

```{r}
dplyr::glimpse(stations)
```

| Column | Type | Description |
|---|---|---|
| `station_name` | character | Station name (title case) |
| `line_number` | integer | Line number |
| `line_name_pt` | character | Line color in Portuguese |
| `line_name` | character | Line color in English |
| `company_name` | character | Operator |
| `type` | character | `"metro"` or `"train"` |
| `status` | character | `"current"` or `"future"` |
| `geometry` | POINT | Station location |

: {.striped .hover}

Transfer stations (e.g., Sé, Paraíso, Ana Rosa) appear once per line they
serve.

# Auxiliary datasets

The package also ships two convenience lookup tables:

- **`metro_lines`** --- maps line numbers to Portuguese and English color
  names (13 rows, covering all lines including planned expansions and the
  network total `line_number = 99`).
- **`metro_colors`** --- named character vector of official hex color codes
  for the six lines with ridership data (e.g., `metro_colors["Blue"]` returns
  `"#171796"`). Useful for consistent plot styling with `scale_color_manual()`.

# Data notes and caveats

## Entrance vs. transported {#entrance-vs-transported}

The METRO source files define these terms as:

- **Entrada de passageiros** (*passenger entries*): passengers entering
  through the turnstile gates (*linha de bloqueios*). This is a station-level
  measurement.
- **Passageiros transportados** (*passengers transported*): the sum of
  turnstile entries **plus** transfer passengers between lines at interchange
  stations (e.g. Sé, Paraíso, Ana Rosa, and Vila Prudente). This is a
  system-level measurement that better captures total demand but
  double-counts passengers who transfer.

The original Portuguese footnote reads:

> Corresponde à soma das entradas pela linha de bloqueios com as
> transferências entre linhas nas estações Sé, Paraíso, Ana Rosa e Vila
> Prudente.

## Station-level transfer counting {#station-transfers}

At interchange stations, the METRO source reports separate figures per line.
For example, at Paraíso (Lines 1 and 2):

- Line 1 figure = passengers boarding Line 1 + transfers from Line 2
- Line 2 figure = passengers boarding Line 2 + transfers from Line 1

This means station-level totals at interchange stations are **not**
double-counted within a single line, but summing across lines at the same
interchange would overcount. The affected stations and their lines are:

| Station | Lines |
|---|---|
| Sé | 1, 3 |
| Paraíso | 1, 2 |
| Ana Rosa | 1, 2 |
| Vila Prudente | 2, 15 |
| Tamanduateí | 2, 10 (CPTM) |

: {.striped .hover}

## Line 5 ownership change {#line5-change}

Line 5 (Lilás) was originally operated by METRO SP. On August 4, 2018, it
was handed over to ViaMobilidade under a concession contract. This affects
the data in two ways:

1. **Source switch**: from October 2017 through July 2018, Line 5 data comes
   from the METRO transparency portal. From August 2018 onward, it comes from
   the Insper Dataverse (ViaMobilidade/Insper partnership).
2. **Transported counts end**: the METRO portal has Line 5 transported data
   through December 2019. The Dataverse does not provide transported counts,
   so `passengers_transported` has no Line 5 data after 2019.

## Station openings during the data window {#station-openings}

Several stations opened during the data period, creating step changes in
line-level totals and partial months at the station level:

| Date | Stations | Line |
|---|---|---|
| 2017-11-27 | Alto da Boa Vista, Borba Gato, Brooklin | 5 |
| 2018-04-02 | Eucaliptos | 5 |
| 2021-01 | Jardim Colonial | 15 |

: {.striped .hover}

A more comprehensive list is available in the `station_inauguration` dataset
(see `?station_inauguration`), though it is still being verified.

## Line 15 Sunday closures

In February and March 2018, Line 15 (Prata) was closed on Sundays for
control system testing. Sunday averages (`mdo`) for these months reflect zero
or near-zero ridership, which is a testing artifact rather than demand.

## Rounding in station averages

The METRO source rounds station-level averages to the nearest thousand. The
sum of individual station values may not equal the line total due to this
rounding. The original note states:

> O total da linha pode ser diferente da soma das estações devido ao
> arredondamento.

## Lines 4 and 5: station codes

The `station_code` column (three-letter abbreviation) is only available for
METRO-operated lines (1, 2, 3, 15). Lines 4 and 5 have `station_code = NA`
because these abbreviations are internal to METRO SP and not used by
ViaQuatro/ViaMobilidade.

## 2017 partial year

Only October through December 2017 is available. The METRO transparency
portal does not provide machine-readable data before October 2017 (earlier
months exist only as PDFs).

## Trailing NA values

The most recent months may have `NA` values for some lines or metrics when
data has not yet been published by the source.

## Source attribution

All METRO portal reports are credited to:

> Gerência de Operações / Coordenadoria de Estratégia Operacional (2017)

or

> Diretoria de Operações / Coordenadoria de Informações Gerenciais e Estudos
> Estratégicos (2018 onward)
