---
title: "Getting Started"
vignette: >
  %\VignetteIndexEntry{getting_started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    fig.width: 8
    fig.asp: 0.618
    fig.align: center
    warning: false
    message: false
---

# metrosp

The `metrosp` package provides access to the [Metro de São Paulo](https://transparencia.metrosp.com.br/) public transportation data. The datasets are compact and the sources publish irregularly, so the package ships the data in a "lazy" format. All data comes prepackaged and is called directly, with no download or import step.

The bundled data is a fixed snapshot, current through July 2026. `read_metro_demand()` reads newer data, published separately on every pipeline run (see [Newer data](#newer-data)).

```{r}
#| label: setup
library(metrosp)
library(dplyr)
```

There are four main datasets:

- `line_entries_monthly`: monthly passengers entering the metro system, by line
- `line_transported_monthly`: monthly passengers transported by the metro system, by line
- `station_entries_monthly`: monthly average of weekday passengers per station
- `station_entries_daily`: daily passengers per station

For convenience, `metrosp` also provides information on stations and lines of the metro system (`rail_lines` and `rail_stations`). The `rail_lines` dataset is a spatial dataset and requires the `sf` package to work properly; it carries the line numbers and Portuguese/English line names for the full network (including planned and CPTM lines).

```{r}
library(sf)

rail_lines
```

The package also provides a named vector of colors for each line of the metro
system (`metro_colors`) and a São Paulo holiday and business-day calendar
(`calendar_spo`).

```{r}
metro_colors
```

Using the datasets is straightforward, just call the dataset name.

```{r}
glimpse(line_entries_monthly)
```

All datasets are returned as `tibble` so using the `dplyr` package is recommended.

## Newer data {#newer-data}

The bundled snapshot stays put across package versions, so examples and analyses remain reproducible. Fresh data goes somewhere else: every pipeline run publishes the four demand datasets to a GitHub release. `read_metro_demand()` reads from there.

```{r}
#| eval: false
# Latest published data
entrance <- read_metro_demand("line_entries_monthly")

# A pinned monthly batch, named in the analysis that used it
entrance_sep <- read_metro_demand("line_entries_monthly", vintage = "2026-09")
```

Downloads use the platform-specific user cache returned by `tools::R_user_dir()`.
`metrosp_cache()` lists its contents and `metrosp_cache_clear()` removes them;
set `cache = FALSE` to keep a download only for the current session. Columns
match the bundled datasets, so code written against one works with the other.

The rest of this tutorial uses the bundled data, which needs no download.

# The datasets

This tutorial will briefly introduce the main datasets and how to use them by making simple visualizations with the data. To better replicate the visualization, use the `ggplot2` package and the custom theme below.

```{r}
#| code-fold: true
library(ggplot2)

theme_series <- theme_minimal(base_family = "Avenir", base_size = 10) +
  theme(
    panel.background = element_rect(fill = "#f5f5f5"),
    plot.background = element_rect(fill = "#f5f5f5"),
    plot.margin = margin(20, 10, 20, 10),
    plot.title = element_text(family = "Lora", size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.25),
    axis.title.x = element_blank(),
    axis.line.x = element_line(color = "gray10", linewidth = 0.5),
    axis.ticks.x = element_line(color = "gray10", linewidth = 0.5),
    strip.background = element_rect(fill = "#0D1B2A"),
    strip.text = element_text(color = "#ffffff"),
    legend.position = "bottom"
  )
```

## Entrance and Transported

Both `line_entries_monthly` and `line_transported_monthly` are monthly series by line. An entry is a passenger crossing a turnstile; a transported passenger is a turnstile entry plus a transfer between lines at an interchange station, so transported counts run above entry counts for the same line and month.

The data is aggregated into metrics:

- `max`: maximum number of passengers (daily peak)
- `mdu`: average number of passengers on business days
- `mdo`: average number of passengers on Sundays
- `msa`: average number of passengers on Saturdays
- `total`: total number of passengers

### Entrance

This dataset is identified by month (`date`), line (`line_number`, `line_name`), and metric (`metric`, `metric_name`). The data is in tidy format and values are in **individual passengers**.

```{r}
glimpse(line_entries_monthly)
```

Because a passenger enters through a turnstile only once, summing `total`, `mdu`, `msa`, or `mdo` across lines gives the corresponding network entry figure. Do not sum `max`: individual lines may peak on different days.

```{r}
total_entrance <- line_entries_monthly |>
  filter(metric == "total")
```

The plot shows the total monthly passenger entrances by metro line. Line 4 starts in 2012 and the other lines in January 2016. July 2017 is the one gap, a month METRO never published an entrance table for.

```{r}
#| code-fold: true
ggplot(total_entrance, aes(x = date, y = value, color = line_name)) +
  geom_line(lwd = 0.8) +
  facet_wrap(vars(line_name), scales = "free_y") +
  scale_color_manual(values = metro_colors) +
  guides(color = "none") +
  labs(
    title = "Total Entrance by Line",
    subtitle = "Total monthly passenger entrances by metro line",
    x = NULL,
    y = "Total Entrance"
  ) +
  theme_series
```

### Transported

This dataset has the same columns as `line_entries_monthly`, but values are in **thousands of passengers**, as METRÔ publishes them. Multiply by 1000 before comparing the two datasets.

Coverage is limited to the METRÔ-operated system. Line 4 never appears, and Line 5 stops in August 2018, when the line passed to ViaMobilidade: the Dataverse source that covers both lines does not publish transported counts.

```{r}
glimpse(line_transported_monthly)
```

A transported passenger is counted on every line used. Summing the lines therefore double-counts interchange journeys and should not be interpreted as the number of unique passengers in the network.

```{r}
daily_avg <- line_transported_monthly |>
  filter(metric == "mdu")
```

The plot below shows the daily average (business days) passengers transported by metro line. The Lilac panel stops in 2018, when Line 5 left the METRÔ reports.

```{r}
#| code-fold: true
ggplot(daily_avg, aes(x = date, y = value, color = line_name)) +
  geom_line(lwd = 0.8) +
  facet_wrap(vars(line_name), scales = "free_y") +
  scale_color_manual(values = metro_colors) +
  labs(
    title = "Daily Average Passenger Transported by Line",
    subtitle = "Monthly averages across business days (thousands)",
    x = NULL,
    y = "Daily Average"
  ) +
  guides(color = "none") +
  theme_series
```

## Station Averages

This dataset is identified by month (`date`), line (`line_number`, `line_name`), physical station (`station_id`), and the constant `mdu` metric. `station_name` is the current display label and `value` is the daily average on business days.

```{r}
glimpse(station_entries_monthly)
```

The plot below shows the daily average (business days) passengers entering each station of line 4. Note that the temporal range of the data is unequal across stations, since not all of them were inaugurated at the same time.

```{r}
#| code-fold: true
line4st <- station_entries_monthly |>
  filter(line_number == 4)

ggplot(line4st, aes(x = date, y = value)) +
  geom_line(lwd = 0.8, color = metro_colors["Yellow"]) +
  facet_wrap(vars(station_name), scales = "free_y") +
  labs(
    x = NULL,
    y = "Average Passengers",
    title = "Passengers per Station (line 4)"
  ) +
  theme_series
```

## Station Daily

This dataset is identified by day (`date`), line (`line_number`, `line_name`), and physical station (`station_id`). `station_name` is the current display label and `value` is the daily number of passengers entering the station. Additionally, `station_code` contains three-letter abbreviations for stations, but only for METRÔ-operated lines.

```{r}
glimpse(station_entries_daily)
```

The plot below shows the trend of daily passengers entering each station of line 4 in 2023.

```{r}
#| code-fold: true
#| fig-width: 10
#| fig-height: 8
line4st_daily <- station_entries_daily |>
  filter(line_number == 4, year == 2023)

ggplot(line4st_daily, aes(x = date, y = value)) +
  geom_smooth(
    lwd = 0.8,
    color = metro_colors["Yellow"],
    method = "loess",
    span = 0.65
  ) +
  facet_wrap(vars(station_name), scales = "free_y", ncol = 3) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    title = "Passengers per Station (line 4, 2023)",
    subtitle = "LOESS smoothed trend",
    x = NULL,
    y = "Average Passengers"
  ) +
  theme_series
```
