---
title: "Getting Started"
vignette: >
  %\VignetteIndexEntry{getting_started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    fig.width: 8
    fig.asp: 0.618
    fig.align: center
    warning: false
    message: false
---

# metrosp

The `metrosp` package provides access to the [Metro de São Paulo](https://www.metrosp.com.br/) public transportation data. Since the data is not updated regularly and the datasets are rather compact, this package distributes the data in a "lazy" format. This means that all data comes prepackaged and is called directly, without needing to download or import the raw data.

```{r}
#| label: setup
library(metrosp)
library(dplyr)
```

There are four main datasets:

- `passengers_entrance`: daily passengers entering the metro system
- `passengers_transported`: daily passengers transported by the metro system
- `station_averages`: daily average passengers per station
- `station_daily`: daily passengers per station

For convenience, `metrosp` also provides information on stations and lines of the metro system (`lines`, `stations`, and `metro_lines`). The `lines` dataset is a spatial dataset and requires the `sf` package to work properly.

```{r}
library(sf)

lines
```

Finally, the package also provides a named vector of colors for each line of the metro system (`metro_colors`).

```{r}
metro_colors
```

## Using the datasets

Using the datasets is straightforward, just call the dataset name.

```{r}
glimpse(passengers_entrance)
```

All datasets are returned as `tibble` so using the `dplyr` package is recommended.

# The datasets

```{r}
library(ggplot2)

theme_series <- theme_minimal(base_family = "Avenir", base_size = 10) +
  theme(
    panel.background = element_rect(fill = "#f5f5f5"),
    plot.background = element_rect(fill = "#f5f5f5"),
    plot.margin = margin(20, 10, 20, 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.25),
    axis.title.x = element_blank(),
    axis.line.x = element_line(color = "gray10", linewidth = 0.5),
    axis.ticks.x = element_line(color = "gray10", linewidth = 0.5),
    strip.background = element_rect(fill = "#0D1B2A"),
    strip.text = element_text(color = "#ffffff"),
    legend.position = "bottom"
  )
```

## Entrance and Transported

The `passengers_entrance` and `passengers_transported` datasets are both monthly passengers entering and transported by the metro system. The former is a daily count of passengers entering the metro system, while the latter is a daily count of passengers transported by the metro system.

The data is aggregated into metrics:

- `max`: maximum number of passengers (daily peak)
- `mdu`: average number of passengers on business days
- `mdo`: average number of passengers on Sundays
- `msa`: average number of passengers on Saturdays
- `total`: total number of passengers

### Entrance

Entrance is calculated by summing the number of passengers entering each station on each day.

```{r}
total_entrance <- passengers_entrance |>
  filter(metric_abb == "total", line_name != "METRO System")

ggplot(total_entrance, aes(x = date, y = value, color = line_name)) +
  geom_line(lwd = 0.8) +
  facet_wrap(vars(line_name), scales = "free_y") +
  scale_color_manual(values = metro_colors) +
  guides(color = "none") +
  labs(x = NULL, y = "Total Entrance") +
  theme_series
```

### Transported

```{r}
daily_avg <- passengers_transported |>
  filter(metric_abb == "mdu", line_number != 99)

ggplot(daily_avg, aes(x = date, y = value, color = line_name)) +
  geom_line(lwd = 0.8) +
  facet_wrap(vars(line_name), scales = "free_y") +
  scale_color_manual(values = metro_colors) +
  labs(x = NULL, y = "Daily Average") +
  theme_series
```

## Station Averages

The `station_averages` dataset is a daily average of passengers per station.

```{r}
line4st <- station_averages |>
  filter(line_number == 4)

ggplot(line4st, aes(x = date, y = avg_passenger)) +
  geom_line(lwd = 0.8, color = metro_colors["Yellow"]) +
  facet_wrap(vars(station_name), scales = "free_y") +
  labs(x = NULL, y = "Average Passengers") +
  theme_series
```

```{r}
top_stations <- station_averages |>
  filter(date == as.Date("2023-08-01"), line_number == 3) |>
  summarise(total = sum(avg_passenger, na.rm = TRUE), .by = station_name) |>
  slice_max(total, n = 10)

top_stations <- top_stations |>
  mutate(
    station_name = factor(
      station_name,
      levels = rev(unique(top_stations$station_name))
    )
  )

ggplot(top_stations, aes(total, station_name)) +
  geom_col(fill = metro_colors["Red"]) +
  scale_fill_manual(values = metro_colors) +
  theme_series
```

```{r}
line4st_daily <- station_daily |>
  filter(line_number == 4, year == 2023)

ggplot(line4st_daily, aes(x = date, y = passengers)) +
  geom_smooth(lwd = 0.8, color = metro_colors["Yellow"], method = "loess", span = 0.65) +
  facet_wrap(vars(station_name), scales = "free_y", ncol = 3) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = NULL, y = "Average Passengers") +
  theme_series
```
